Global sensitivity analysis of model uncertainty in aeroelastic wind turbine models

### 2020-06-02 Papers/Torque_Paper/revision/content.tex
added Torque revision; updated folder structure

\section{Global sensitivity analysis}\label{sec:GSA}
The objective of sensitivity analysis is to quantify the relative significance of individual inputs (or their combination) and how variations in input values affect the output of interest. Sensitivity analysis techniques can be classified as local and global \cite{RSmith}. In a local sensitivity analysis, individual parameters are perturbed around their nominal values allowing for the description of output variability only in a small neighbourhood around the nominal input values. Although local approaches are widely employed due to their ease of implementation and low computational cost, they are unable to quantify the global behaviour of nonlinearly parametrized models such as aeroelastic models. Global sensitivity approaches, on the other hand, consider the entire range of input values to compute output sensitivities. Therefore, a global sensitivity analysis is more suitable for the aeroelastic models considered in this work.  
%In engineering, sensitivity analysis can be employed for several reasons: to determine the stability and robustness of a computational model for input parameters, for simplification of stochastic models by fixing the insensitive parameters, and to guide data acquisition campaigns and experimental design to refine the data on sensitive parameters.

\subsection{Sobol analysis}
We employ the variance-based Sobol decomposition to perform global sensitivity analysis. This approach allows for quantification of the relative importance of the input parameters known as \emph{Sobol indices}. In a Sobol analysis, we express the total variance of the output in term of contributions from individual parameters and their combinations. Consider a generic BEM model $f(\mathbf{z})$, depending on uncertain parameters $\mathbf{z}$ and returning a quantity of interest $Y$ (e.g.\ power, loads, moments):
\begin{equation}\label{nonlinear_model}
Y = f({\mathbf{z}}),
\end{equation}
where ${\mathbf{z}} = [z_1, z_2, ..., z_M]\in \mathcal{D}_{\mathbf{z}}\in \mathbb{R}^M$ is a vector of random variables (parameters). For ease of notation, we assume these input parameters to be uniformly distributed i.e. $z_i \sim \mathcal{U}(0,1)$ and the support of the input set is $\mathcal{D}_{\mathbf{z}} = [0,1]^M$ where $M$ is the total number of input parameters. The Sobol decomposition is based on the following hierarchical representation \cite{RSmith}:
\begin{equation}\label{sobol_decomp}
f(z_1, z_2, ..., z_M) = f_0+\sum_{i=1}^M f_i(z_i) + \sum_{1\leq i<j\leq M} f_{ij}(z_i,z_j) + ... + f_{1, 2, ..., M}(z_1, z_2, ..., z_M).
\end{equation}
Here the zeroth-order function $f_0$ is the mean response of $f$, the first-order univariate functions $f_i(z_i)$ quantify independent contributions due to each random input, and second-order functions $f_{i,j}(z_i,z_j)$ represent the effect of interaction between $z_i$ and $z_j$ on the response $f$. Formally, the zeroth-, first- and second-order terms are defined as:
\begin{align}\label{sobol_terms}
f_0 = \int_{\mathcal{D}_{\mathbf{z}}}f(\mathbf{z})d\mathbf{z}, \qquad  f_i(z_i)= \int_{{\mathcal{D}_{\mathbf{z}}}^{M-1}}f(\mathbf{z})d\mathbf{z}_{\sim \{i\}} - f_0,\\
f_{ij}(z_i,z_j) = \int_{{\mathcal{D}_{\mathbf{z}}}^{M-2}}f(\mathbf{z})d\mathbf{z}_{\sim \{i,j\}} - f_i(z_i) - f_j(z_j) - f_0,
\end{align}
where ${\mathcal{D}_{\mathbf{z}}}^{M-1} = [0,1]^{M-1}$ and the notation $\mathbf{z}_{\sim \{i,j\}}$ denotes the vector having all the components of $\mathbf{z}$ except in the set $\{i,j\}$. Higher-order terms are interpreted in a similar manner. Note that \eqref{sobol_decomp} is only valid for independent input parameters. We consider a second-order expansion of \eqref{sobol_decomp}, which is generally sufficient for the purpose of this article.

@@

Sobol indices are based on a variance decomposition \cite{SOBOL2001271}. The definition of the variance is:
\begin{equation}\label{tot_var}
D = \text{Var}[f(\mathbf{z})] = \int_{\mathcal{D}_{\mathbf{z}}} f^2(\mathbf{z})d\mathbf{z} - f_0^2,
\end{equation}
which is rewritten as
%As the expansion in \eqref{sobol_decomp} is not unique, some orthogonality conditions must be imposed  \cite{Rabitz1999,} which allows to express the total variance as
\begin{equation}
D = \sum_{i=1}^M D_i + \sum_{1\leq i<j\leq M} D_{ij},
\end{equation}
where the first- and second-order partial variances are defined via decomposition \eqref{sobol_decomp} as
\begin{align}
D_i &= \int_{0}^{1} f^2_i(z_i)dz_i, & D_{ij} &= \int_{0}^{1} \int_{0}^{1} f^2_{ij}(z_i,z_j)dz_idz_j \label{partial_variance2}. 
\end{align}
The first- and second-order Sobol indices are then computed as
\begin{equation}\label{sobol_ind}
S_i = \frac{D_i}{D}, \quad S_{ij} = \frac{D_{ij}}{D}, \quad i,j=1,2, ..., M.
\end{equation}
%and the sum of all indices satisfy
%\begin{equation}
%\sum_{i=1}^M S_i + \sum_{1\leq i<j\leq M} S_{ij} = 1.
%\end{equation}
The total effect of $z_i$ on the output $Y$ can be quantified using the total sensitivity index $S_{T_i} = S_i + \sum_{j=1}^M S_{ij}$. The total sensitivity indices can be interpreted as an importance measure for the parameter $z_i$, therefore a large $S_{T_i}$ implies that $z_i$ has a strong influence on $Y$. The computation of $S_{T_i}$ requires the approximation of partial variances defined in \eqref{partial_variance2}. Using the Monte Carlo method for computing these variances can be prohibitive for computationally expensive models $Y$. On the other hand, Polynomial Chaos Expansions (PCE) based approaches provide a more efficient alternative for sensitivity analysis, and will be used here. As shown in \cite{SUDRET2008964}, one can compute the Sobol indices analytically by post-processing the PCE coefficients, whose computation will be detailed next.

\subsection{PCE-based Sobol indices computation}
A PCE approximation $f^{K}(\mathbf{z})$ to the computational model $Y=f(\mathbf{z})$ can be defined as a weighted sum of multivariate polynomials in $\mathbf{z}$ \cite{RSmith}:
\begin{equation}\label{eq:PCE}
Y= f(\mathbf{z}) \approx f^{K}(\mathbf{z}) = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}),
\end{equation}
where $\mathbf{k}\in \mathbb{N}_0^M$ is multi-index with magnitude $|\mathbf{k}| = k_1+k_2 + ... + k_M$ and $\Psi_{\mathbf{k}}(\mathbf{z})$ a multivariate polynomial basis, computed using a tensor product of univariate polynomials $\psi_{k_i}^{(i)}(z_i)$, 
%\begin{equation}
%\Psi_{\mathbf{k}}(\mathbf{z}) := \prod_{i=1}^M\psi_{k_i}^{(i)}(z_i),
%\end{equation}
where $k_i$ is the order of the univariate polynomial. The choice of univariate orthogonal polynomial $\psi_{k_i}$ depends on the type of the random variable $z_i$. For example, for uniform random variables the Legendre family of polynomials is used. 
%The multivariate polynomials $\Psi_{\mathbf{k}}(\mathbf{z})$ is orthogonal with respect to the joint probability distribution of random vector $\mathbf{z}$. 
The total number of terms $K$ in \eqref{eq:PCE} depends on the truncation scheme; in a standard truncation scheme  all polynomials in $M$ random variable with order less than or equal to $p$ are used. In this work, a more advanced truncation scheme known as the hyperbolic truncation scheme is used \cite{BlatmanThesis}. 
%\textcolor{red}{In this work, we use the hyperbolic truncation scheme with a $qnorm = 0.75$.}

The PCE coefficients $w_{\mathbf{k}}$ corresponding to $\Psi_{\mathbf{k}}(\mathbf{z})$ can be computed using a pseudo-spectral approach by using quadrature methods. For a small number of dimensions this is an efficient approach. However, in the current work we will encounter more than 10 dimensions and quadrature methods become  too expensive as they suffer strongly from the curse of dimensionality. We therefore focus on methods in which the sampling points can be chosen more freely and adaptively: Ordinary Least Squares (OLS) and Least-Angle Regression (LARS).
%In the current work where we have a limited computational budget, methods that minimizes the approximation error for a given number of samples are more practical. Next, we briefly describe two such methods to compute PCE coefficients based on a fixed number of model evaluations.

@@

\subsubsection{Ordinary Least Squares}
%Instead of using a projection-based method one can also use least squares based approaches to compute the coefficients \cite{Berveiller2006}. 
The PCE expansion \eqref{eq:PCE} can be expressed in terms of the following vector-matrix product
\begin{equation}\label{eq:LSM}
Y = f^{K}(\mathbf{z}) +\varepsilon_K = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}) + \varepsilon_K = \mathbf{w}^T\boldsymbol{\Psi}(\mathbf{z}) +\varepsilon_K,
\end{equation} 
where $\mathbf{w} = \{w_0,w_1,...,w_K\}^T$ is the coefficient vector, $\boldsymbol{\Psi}(\mathbf{z}) = \{\Psi_0(\mathbf{z}), \Psi_1(\mathbf{z}), ..., \Psi_K(\mathbf{z})\}^T$ is the matrix containing multivariate polynomials in $\mathbf{z}$, and $\varepsilon_K$ is the approximation error. The least-squares minimization problem can then be formulated as:
\begin{equation}\label{eq:PCE_OLS}
\mathbf{\overline{w}} = \argmin_{\mathbf{w}} \mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right].
 \end{equation}
The coefficient vector $\mathbf{\overline{w}}$ can be computed using the Ordinary Least Squares (OLS) method. Given $N$ model evaluations $\mathbf{Y} = \{f(\mathbf{z}_1), f(\mathbf{z}_2), ..., f(\mathbf{z}_N)\}$, we can obtain the coefficients as $\overline{\mathbf{w}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{Y}$, where $\mathbf{X}\in \mathbb{R}^{N\times K}$ is the design matrix where $\mathbf{X}_{i,j} = \Psi_{j}(\mathbf{z}_i)$, $i = 1,2, ..., N$ and $j = 0,1,2,..., K$. Although the OLS method allows to compute the coefficients using a limited number of model evaluations, it may result in a PCE model that consists of high-order interaction terms that may lead to overfitting. For many engineering problems, low-order interactions between the input variables are most important. Therefore, the least squares form in \eqref{eq:PCE_OLS} can be regularized such that the minimization results in a low-order sparse PCE model, as discussed next.

\subsubsection{Least Angle Regression}
The LARS algorithm \cite{efron2004} is one of the methods that can be applied to compute sparse PCE models \cite{BlatmanThesis}. The key idea is to find the best set of polynomials from a given candidate set that minimizes the regularized form of \eqref{eq:PCE_OLS}:
\begin{equation}\label{regularized_LS}
\mathbf{\overline{w}} = \argmin_{\mathbf{w}} \mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right] + \lambda ||\mathbf{w}||_1.
\end{equation}
The LARS algorithm in the context of PCE can be summarized as follows. We start with all PCE coefficients for the candidate polynomials set to zero, and find the polynomial, say $\Psi_1(\mathbf{w})$ most correlated with the model evaluations. This polynomial enters the active set. Next, we take the largest step in the direction of $\Psi_1(\mathbf{w})$ until another polynomial $\Psi_2(\mathbf{w})$ from the candidate set, has as much correlation with the current residual. The second polynomial is now also added to the active set. In the next step, we move both the coefficients in the active set towards their least-square values until a third polynomial from the candidate set shows as much correlation as the current PCE model based on two polynomials shows with the model evaluations. The number of iterations is given by the minimum of $\{K,N\}$. After every iteration an a-posteriori error, for instance the Leave-One-Out (LOO) cross-validation error, is computed. The active set with the smallest LOO is then chosen as the best sparse PCE model. One of the main advantage of the LARS algorithm is that it works well in cases when $K>>N$, i.e. the number of polynomials in the candidate set is much larger than the number of model evaluations.

@@

\subsection{Choice of uncertain inputs}
The data for lift ($C_l$) and drag ($C_d$) polars are available at four locations along the blade radius: at 11.87 m, 17.82 m, 28.97 m, and 35.53 m. The lift and drag variables at these sections are numbered Cl1 - Cl4 and Cd1 - Cd4 respectively. The reference value of these polars is obtained from the wind-tunnel experiment with 3D corrections. Random samples of chord, twist, lift- and drag-polars are obtained by perturbing the control points with a uniformly distributed random variable.

There are two ways to obtain perturbed curves: independently perturbing each of the control points (local perturbation) or using a single random number to perturb all the control points (global perturbation). We use local perturbation for chord control points Ch3 - Ch7 and twist control points Tw2 - Tw7 control points, see Fig.\ \ref{perturbed_samples} (a) - (b), respectively. The twist and chord are perturbed following equation \eqref{eqn:perturb_c}, whereas the lift and drag are perturbed according to equation \eqref{eqn:perturb_cl}. Examples of the resulting lift and drag curves are shown in Fig.\ \ref{perturbed_samples} (c) - (d), respectively. For simplicity, we only consider uniform distributions for perturbing the control points, however, we can also use other probability distributions for this purpose. Also note that when using a uniform distribution with bounds given by the relative magnitude of control points, we tend to obtain small perturbations for control points with values close to zero. 
\begin{figure}[ht!]
\centering
\includegraphics[trim={0.5cm 0cm 0 0cm},clip, scale=0.67]{figure1.pdf}
\caption{Random realization of chord, twist, lift- (Cl2) and drag-polars (Cd2).}
\label{perturbed_samples}
\end{figure}

\subsection{Verification of sensitivity analysis method}
We perform a convergence study for both OLS and LARS in order to determine which method converges fastest with increasing number of uncertain parameters and number of model runs. For this purpose we only include uncertainty in the chord and set the power production as the quantity of interest. The NURBS representation is defined by 9 control points, of which a subset is taken uncertain (see Fig.\ \ref{perturbed_samples}(a)). In Fig.\ \ref{convergence_dim}, we show for both OLS and LARS the convergence behaviour of the total Sobol indices of the power output with increasing number of model evaluations $N$ (runs of the AeroModule), and with increasing number of uncertain inputs $M$ (4, 6 and 8 control points). We see that both methods scale well with increasing dimensions, however, LARS exhibits faster convergence than OLS in high dimensions due to its adaptive nature, and will be the method of choice in the subsequent tests. For both OLS and LARS, we start with a small $N$ and double it with every iteration until converged values of Sobol indices are obtained.
%}  \todo{what is a reasonable choice of $N$, depending on $M$?}

@@

\subsection{Sensitivity analysis results}
In Fig.~\ref{fig:sensitivity_analysis_power} and \ref{fig:sensitivity_analysis_force}, we show the total order Sobol indices as a measure of the sensitivity of the different geometric and model parameters on the power output and the axial force, respectively, computed using LARS with $M=19$ uncertain parameters chosen from Fig.~\ref{perturbed_samples}. We found that $N=1024$ model evaluations were sufficient to achieve converged indices. In this case, we observe that the uncertainty in model parameters is at least as important as the uncertainty in geometric parameters. Furthermore, within the geometric parameters, the chord distribution shows a significantly higher sensitivity than the twist variables. Lastly, we observe that the axial force is basically only sensitive to the lift coefficient at the outward blade sections, whereas the power is sensitive to both lift and drag coefficients, also at the inward section of the blade. We note that these findings are testcase-specific and might not generalize directly to other turbines or other operational settings. Furthermore, the reported sensitivities depend on the variance that is chosen for the uncertain parameters; in future work we will relate these to values known from design standards.

>>>>

\section{Global sensitivity analysis}\label{sec:GSA}
The objective of sensitivity analysis is to quantify the relative significance of individual inputs (or their combination) and how variations in input values affect the output of interest. Sensitivity analysis techniques can be classified as local and global \cite{RSmith}. In a local sensitivity analysis, individual parameters are perturbed around their nominal values allowing for the description of output variability only in a small neighbourhood around the nominal input values. Although local approaches are widely employed due to their ease of implementation and low computational cost, they are unable to quantify the global behaviour of nonlinearly parametrized models such as aeroelastic models. Global sensitivity approaches, on the other hand, consider the entire range of input values to compute output sensitivities, and will be employed in this work. \hl{As this technique is still relatively new in the wind engineering community, we will describe it in some detail in the following sections.}
%In engineering, sensitivity analysis can be employed for several reasons: to determine the stability and robustness of a computational model for input parameters, for simplification of stochastic models by fixing the insensitive parameters, and to guide data acquisition campaigns and experimental design to refine the data on sensitive parameters.

\subsection{Sobol analysis}
We employ the variance-based Sobol decomposition to perform global sensitivity analysis and quantify the relative importance of input parameters via \emph{Sobol indices}. The total variance of the output is expressed in term of contributions from individual parameters and their combinations. Consider a generic BEM model $f(\mathbf{z})$, depending on uncertain parameters $\mathbf{z}$ and returning a quantity of interest $Y$ (e.g.\ power, loads, moments):
\begin{equation}\label{nonlinear_model}
Y = f({\mathbf{z}}),
\end{equation}
where ${\mathbf{z}} = [z_1, z_2, ..., z_M]\in \mathcal{D}_{\mathbf{z}}\in \mathbb{R}^M$ is a vector of random variables (parameters). For ease of notation, we assume these input parameters to be uniformly distributed, i.e.\ $z_i \sim \mathcal{U}(0,1)$, and $\mathcal{D}_{\mathbf{z}} = [0,1]^M$ where $M$ is the total number of input parameters. The Sobol decomposition is based on the following representation \cite{RSmith}:
\begin{equation}\label{sobol_decomp}
f(z_1, z_2, ..., z_M) = f_0+\sum_{i=1}^M f_i(z_i) + \sum_{1\leq i<j\leq M} f_{ij}(z_i,z_j) + ... + f_{1, 2, ..., M}(z_1, z_2, ..., z_M).
\end{equation}
The zeroth-order function $f_0$ is the mean response of $f$, the first-order univariate functions $f_i(z_i)$ quantify independent contributions due to each random input, and second-order functions $f_{i,j}(z_i,z_j)$ represent the effect of interaction between $z_i$ and $z_j$ on the response $f$:
\begin{align}\label{sobol_terms}
f_0 = \int_{\mathcal{D}_{\mathbf{z}}}f(\mathbf{z})d\mathbf{z}, \qquad  f_i(z_i)= \int_{{\mathcal{D}_{\mathbf{z}}}^{M-1}}f(\mathbf{z})d\mathbf{z}_{\sim \{i\}} - f_0,\\
f_{ij}(z_i,z_j) = \int_{{\mathcal{D}_{\mathbf{z}}}^{M-2}}f(\mathbf{z})d\mathbf{z}_{\sim \{i,j\}} - f_i(z_i) - f_j(z_j) - f_0,
\end{align}
where the notation $\mathbf{z}_{\sim \{i,j\}}$ denotes the vector having all the components of $\mathbf{z}$ except in the set $\{i,j\}$. Higher-order terms are interpreted in a similar manner. Note that \eqref{sobol_decomp} is only valid for independent input parameters. We consider a second-order expansion of \eqref{sobol_decomp}, which is generally sufficient for the purpose of this article.

@@

Sobol indices are based on the decompositon of the variance \cite{SOBOL2001271}, defined as:
\begin{equation}\label{tot_var}
D = \text{Var}[f(\mathbf{z})] = \int_{\mathcal{D}_{\mathbf{z}}} f^2(\mathbf{z})d\mathbf{z} - f_0^2,
\end{equation}
which is rewritten as
%As the expansion in \eqref{sobol_decomp} is not unique, some orthogonality conditions must be imposed  \cite{Rabitz1999,} which allows to express the total variance as
\begin{equation}
D = \sum_{i=1}^M D_i + \sum_{1\leq i<j\leq M} D_{ij},
\end{equation}
where the first- and second-order partial variances are defined via decomposition \eqref{sobol_decomp} as
\begin{align}
D_i &= \int_{0}^{1} f^2_i(z_i)dz_i, & D_{ij} &= \int_{0}^{1} \int_{0}^{1} f^2_{ij}(z_i,z_j)dz_idz_j \label{partial_variance2}. 
\end{align}
The first- and second-order Sobol indices are then computed as
\begin{equation}\label{sobol_ind}
S_i = \frac{D_i}{D}, \quad S_{ij} = \frac{D_{ij}}{D}, \quad i,j=1,2, ..., M.
\end{equation}
%and the sum of all indices satisfy
%\begin{equation}
%\sum_{i=1}^M S_i + \sum_{1\leq i<j\leq M} S_{ij} = 1.
%\end{equation}
The total effect of $z_i$ on the output $Y$ can be quantified using the total sensitivity index $S_{T_i} = S_i + \sum_{j=1}^M S_{ij}$. The total sensitivity indices can be interpreted as an importance measure for the parameter $z_i$: a large $S_{T_i}$ implies that $z_i$ has a strong influence on $Y$. The computation of $S_{T_i}$ requires the approximation of partial variances defined in \eqref{partial_variance2}. Using the Monte Carlo method for computing these variances is computationally very expensive, and we therefore use instead a Polynomial Chaos Expansions (PCE) based approach. As shown in \cite{SUDRET2008964}, one can compute the Sobol indices analytically by post-processing the PCE coefficients.

\subsection{PCE-based Sobol indices computation}
A PCE approximation $f^{K}(\mathbf{z})$ to the computational model $Y=f(\mathbf{z})$ can be defined as a weighted sum of multivariate polynomials in $\mathbf{z}$ \cite{RSmith}:
\begin{equation}\label{eq:PCE}
Y= f(\mathbf{z}) \approx f^{K}(\mathbf{z}) = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}),
\end{equation}
where $\mathbf{k}\in \mathbb{N}_0^M$ is multi-index with magnitude $|\mathbf{k}| = k_1+k_2 + ... + k_M$ and $\Psi_{\mathbf{k}}(\mathbf{z})$ a multivariate polynomial basis, computed using a tensor product of univariate polynomials $\psi_{k_i}^{(i)}(z_i)$, 
%\begin{equation}
%\Psi_{\mathbf{k}}(\mathbf{z}) := \prod_{i=1}^M\psi_{k_i}^{(i)}(z_i),
%\end{equation}
where $k_i$ is the order of the univariate polynomial. The choice of univariate orthogonal polynomial $\psi_{k_i}$ depends on the type of the random variable $z_i$. For the uniform random variables appearing in this work Legendre polynomials are employed. 
%The multivariate polynomials $\Psi_{\mathbf{k}}(\mathbf{z})$ is orthogonal with respect to the joint probability distribution of random vector $\mathbf{z}$. 
The total number of terms $K$ in \eqref{eq:PCE} depends on the truncation scheme; in this work, a hyperbolic truncation scheme is used \cite{BlatmanThesis}. The PCE coefficients $w_{\mathbf{k}}$ corresponding to $\Psi_{\mathbf{k}}(\mathbf{z})$ can be computed using a pseudo-spectral approach by using quadrature methods. For a small number of dimensions this is an efficient approach. However, in the current work we will encounter more than 10 dimensions and quadrature methods become too expensive as they suffer strongly from the curse of dimensionality. We therefore focus on methods in which the sampling points can be chosen more freely and adaptively: Ordinary Least Squares (OLS) and Least-Angle Regression (LARS).
%In the current work where we have a limited computational budget, methods that minimizes the approximation error for a given number of samples are more practical. Next, we briefly describe two such methods to compute PCE coefficients based on a fixed number of model evaluations.

@@

\subsubsection{Ordinary Least Squares}
%Instead of using a projection-based method one can also use least squares based approaches to compute the coefficients \cite{Berveiller2006}. 
The PCE expansion \eqref{eq:PCE} can be expressed in terms of the following vector-matrix product
\begin{equation}\label{eq:LSM}
Y = f^{K}(\mathbf{z}) +\varepsilon_K = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}) + \varepsilon_K = \mathbf{w}^T\boldsymbol{\Psi}(\mathbf{z}) +\varepsilon_K,
\end{equation} 
where $\mathbf{w} = \{w_0,w_1,...,w_K\}^T$ is the coefficient vector, $\boldsymbol{\Psi}(\mathbf{z}) = \{\Psi_0(\mathbf{z}), \Psi_1(\mathbf{z}), ..., \Psi_K(\mathbf{z})\}^T$ is the matrix containing multivariate polynomials in $\mathbf{z}$, and $\varepsilon_K$ is the approximation error. The least-squares minimization problem can then be formulated as:
\begin{equation}\label{eq:PCE_OLS}
\mathbf{\overline{w}} = \argmin_{\mathbf{w}} \mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right].
 \end{equation}
The coefficient vector $\mathbf{\overline{w}}$ can be computed using the Ordinary Least Squares (OLS) method. Given $N$ model evaluations $\mathbf{Y} = \{f(\mathbf{z}_1), f(\mathbf{z}_2), ..., f(\mathbf{z}_N)\}$, we can obtain the coefficients as $\overline{\mathbf{w}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{Y}$, where $\mathbf{X}\in \mathbb{R}^{N\times K}$ is the design matrix where $\mathbf{X}_{i,j} = \Psi_{j}(\mathbf{z}_i)$, $i = 1,2, ..., N$ and $j = 0,1,2,..., K$. Although the OLS method allows to compute the coefficients using a limited number of model evaluations, it may result in a PCE model that consists of high-order interaction terms leading to overfitting. For many engineering problems, low-order interactions between the input variables are most important. Therefore, the least squares form in \eqref{eq:PCE_OLS} can be regularized such that the minimization results in a low-order sparse PCE model, as discussed next.

\subsubsection{Least Angle Regression}
The LARS algorithm \cite{efron2004} is one of the methods that can be applied to compute sparse PCE models \cite{BlatmanThesis}. The key idea is to find the best set of polynomials from a given candidate set that minimizes the regularized form of \eqref{eq:PCE_OLS}:
\begin{equation}\label{regularized_LS}
\mathbf{\overline{w}} = \argmin_{\mathbf{w}} \mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right] + \lambda ||\mathbf{w}||_1.
\end{equation}
The LARS algorithm in the context of PCE starts with all PCE coefficients set to zero, and then iteratively selects polynomials based on the correlation with the current residual. After every iteration an a-posteriori error, for instance the Leave-One-Out (LOO) cross-validation error, is computed. The active set with the smallest LOO is then chosen as the best sparse PCE model. One of the main advantages of the LARS algorithm is that it works well in cases when the number of polynomials in the candidate set is much larger than the number of model evaluations.

%The LARS algorithm in the context of PCE can be summarized as follows. We start with all PCE coefficients for the candidate polynomials set to zero, and find the polynomial, say $\Psi_1(\mathbf{w})$ most correlated with the model evaluations. This polynomial enters the active set. Next, we take the largest step in the direction of $\Psi_1(\mathbf{w})$ until another polynomial $\Psi_2(\mathbf{w})$ from the candidate set, has as much correlation with the current residual. The second polynomial is now also added to the active set. In the next step, we move both the coefficients in the active set towards their least-square values until a third polynomial from the candidate set shows as much correlation as the current PCE model based on two polynomials shows with the model evaluations. The number of iterations is given by the minimum of $\{K,N\}$. After every iteration an a-posteriori error, for instance the Leave-One-Out (LOO) cross-validation error, is computed. The active set with the smallest LOO is then chosen as the best sparse PCE model. One of the main advantage of the LARS algorithm is that it works well in cases when $K>>N$, i.e. the number of polynomials in the candidate set is much larger than the number of model evaluations.

@@

\subsection{Choice of uncertain inputs}
The data for lift ($C_l$) and drag ($C_d$) polars are available at four locations along the blade radius: at 11.87 m, 17.82 m, 28.97 m, and 35.53 m. The lift and drag variables at these sections are numbered Cl1 - Cl4 and Cd1 - Cd4 respectively. The reference value of these polars is obtained from the wind-tunnel experiment with 3D corrections. Random samples of chord, twist, lift- and drag-polars are obtained by perturbing the control points with a uniformly distributed random variable \hl{based on Latin Hypercube Sampling}.

There are two ways to obtain perturbed curves: independently perturbing each of the control points (local perturbation) or using a single random number to perturb all the control points (global perturbation). We use local perturbation for chord control points Ch3 - Ch7 and twist control points Tw2 - Tw7 control points, see Fig.\ \ref{perturbed_samples} (a) - (b), respectively. The twist and chord are perturbed following equation \eqref{eqn:perturb_c}, whereas the lift and drag are perturbed according to equation \eqref{eqn:perturb_cl}. \hl{Note that the multiplication of $(1+\Delta_i)$ with the unperturbed value (equations \eqref{eqn:perturb_c}, \eqref{eqn:perturb_cl}) has some physical intuition (especially for the lift and drag curves), but causes relatively large input uncertainty for the chord at the inboard sections. Similarly, we tend to obtain small perturbations for control points with values close to zero.} Examples of the resulting lift and drag curves are shown in Fig.\ \ref{perturbed_samples} (c) - (d), respectively. For simplicity, we only consider uniform distributions for perturbing the control points, however, we can also use other probability distributions for this purpose. 
\begin{figure}[ht!]
\centering
\includegraphics[trim={0.5cm 0cm 0 0cm},clip, scale=0.67]{figure1.pdf}
\caption{Random realization of chord, twist, lift- (Cl2) and drag-polars (Cd2).}
\label{perturbed_samples}
\end{figure}

\subsection{Verification of sensitivity analysis method}
We perform a convergence study for both OLS and LARS in order to determine which method converges fastest with increasing number of uncertain parameters and number of model runs. For this purpose we only include uncertainty in the chord and set the power production as the quantity of interest. The NURBS representation is defined by 9 control points, of which a subset is taken uncertain (see Fig.\ \ref{perturbed_samples}(a)). In Fig.\ \ref{convergence_dim}, we show for both OLS and LARS the convergence behaviour of the total Sobol indices of the power output with increasing number of model evaluations $N$ (runs of the AeroModule), and with increasing number of uncertain inputs $M$ (4, 6 and 8 control points). We see that both methods scale well with increasing dimensions, however, LARS exhibits faster convergence than OLS in high dimensions due to its adaptive nature. \hl{This is consistent with results from literature \cite{uqlab}, and therefore LARS will be the method of choice in the subsequent tests.} For both OLS and LARS, we start with a small $N$ and double it with every iteration until converged values of Sobol indices are obtained.
%}  \todo{what is a reasonable choice of $N$, depending on $M$?}

@@

\subsection{Sensitivity analysis results}
In Fig.~\ref{fig:sensitivity_analysis_power} and \ref{fig:sensitivity_analysis_force}, we show the total order Sobol indices as a measure of the sensitivity of the different geometric and model parameters on the power output and the axial force, respectively, computed using LARS with $M=19$ uncertain parameters chosen from Fig.~\ref{perturbed_samples}. We found that $N=1024$ model evaluations were sufficient to achieve converged indices. In this case, we observe that the uncertainty in model parameters is at least as important as the uncertainty in geometric parameters. Furthermore, within the geometric parameters, the chord distribution shows a significantly higher sensitivity than the twist variables. Lastly, we observe that the axial force is basically only sensitive to the lift coefficient at the outward blade sections, whereas the power is sensitive to both lift and drag coefficients, also at the inward section of the blade. We note that these findings are testcase-specific and might not generalize directly to other turbines or other operational settings. Furthermore, the reported sensitivities depend on the variance that is chosen for the uncertain parameters. For example, \hl{the large sensitivity of the power with respect to uncertainty in Ch2 as shown in figure \ref{fig:sensitivity_analysis_power} is mainly due to the fact that the effective input uncertainty is larger here than at the outboard stations because of the multiplication with the local chord value (see equation \eqref{eqn:perturb_c}). Similarly, the low sensitivity with respect to twist as apparent in figures \ref{fig:sensitivity_analysis_power}-\ref{fig:sensitivity_analysis_force} is due to the fact that the effective perturbation in the twist is rather small: around 0.5 degree at the mid-span location (see equation \eqref{eqn:perturb_c}).} In future work we will relate these perturbations to values known from design standards to arrive at more realistic predictions.
